Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-09 inc/photocathode/plasmon/plasmon.tex
merging comments from andreas on PAPE section

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this paper, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
%TODO should this be formalized, either one way (SPP) or the other
Electrons in a material can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the material creates strong oscillatory local surface electric field \cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.

This avenue of investigation was motivated by Zawadzka et al \cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors mount a thin gold foil on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the ``grating coupling'' geometry.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this geometry the back surface is now free to be attached to a heat sink.

@@

Every material has an intrinsic plasma frequency given by 
\begin{equation}
  \omega_{P} = \sqrt{\frac{\rho_{e} e^2}{\varepsilon_{0} m^*}} \,\text{,}
\end{equation}
where $\rho_{e}$ is the electron density and $m*$ is the effective mass of the material.
And in the most simple (undamped) approximations, the dielectric function of the material may be written in terms of this constant
\begin{equation} \label{eq:dielectric_as_omega}
  \varepsilon(\omega) = 1 - \frac{ \omega_{P}^2 }{ \omega^2 } \,\text{.}
\end{equation}

At the interface of a vacuum and such a simple material material, contstraints on the material parameters which may allow surface plasmons result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $.
Substituting this result into \ref{eq:dielectric_as_omega} show that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons carry momentum whose magnitude is given by
\begin{equation}
  k_{P}^2 = \frac{\omega^2}{c^2} \left( \frac{\varepsilon_{1} \varepsilon_{2}}{\varepsilon_{1} + \varepsilon_{2}} \right) \,\text{,}
\end{equation}
but of course for a vacuum interface this is simplified to 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

In the grating coupling scheme (shown in \ref{fig:plasmon_schematic}) a laser of wavelength $\lambda$ is incident on a material with a grating of period $d$, making an angle $\theta$ from the surface normal. A surface plasmon is driven when the portion of the laser wave-vector parallel to the plasmon ($k_{Lx}$) and a fourier component ($k_{G} = 2 \pi / d$) of the grating matches the wave-vector of the surface plasmon
\begin{equation}
  k_{SP} = k_{Lx} + n k_{G} \,\text{,}
\end{equation}
where $n$ here is an integer.
Which reduces to 
\begin{equation} \label{eq:plasmon_condition}
  \sqrt{ \frac{ \varepsilon }{ 1 + \varepsilon } } = \sin \theta + \frac{ n \lambda }{ d } \,\text{.}
\end{equation}
For a given grating period, \ref{eq:plasmon_condition} determines the angle of incidence that the laser must make with the surface in order to drive the plasmon.
Further, it is used to design the grating, allowing for angles that the microscope column will accept.

@@

For proof-of-concept purposes the first experiment was run on a commercially available gold foil on glass holographic grating (750 lines/mm).
Like the Kretschmann geometry, a foil on glass grating is vulnerable to heating problems; a plasmon creates a large amount of heat but the glass is a poor heat conductor.
The aforementioned laser (emitting $\sim$200fs pulses of green light at 30MW/cm$^2$) was used to drive the plasmon.

The input laser angle was slowly changed until the plasmon oscillation is visible.
At the instant that the plasmon started, a bright flash was visible on the scintillator screen then disappeared.
After this, the grating showed visual signs of damage, both on the surface and in the reflected light pattern.
\ref{fig:grating-damage} shows the results of Scanning Electron Microscopy (SEM) on the grating.
The damage pattern is round and of the size of the laser spot used to drive the plasmon.
The extent of the damage indicates a large temperature increase, which given the other evidence of the test, can be assumed to be from the sudden appearance of a surface plasmon.

There are several conclusions that can be drawn from this experiment.
Most importantly, it shows that a plasmon was driven successfully given the angular dependence and large response, few other processes could have occurred.
The heat generated by the plasmon oscillation destroyed the grating, therefore it validated the decision not to pursue a Kretschmann geometry over fears of heat damage at these high powers.
Finally, it suggests that if using a thermally conductive and structurally sound photoemission material and mount for the plasmonic photocathode, PAPE might generate a large number of electrons (as evident by the bright flash).

@@

@@

After the destruction of the gold foil on glass grating, new gratings were designed to have a higher thermal conductivity.
The second generation grating was constructed from a silicon substract then coated with a gold film.
The thermal conductivity of silicon is approximately 100 times that of glass. %TODO reference needed?

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to iteratively mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$ \cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.

@@

A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is several times thicker than the penetration depth of green light on gold, the laser radiation cannot interact with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?

>>>>

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a conducting material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this thesis, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
Electrons in a conductor can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the conductor creates strong oscillatory local surface electric field \cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.
On the other hand, these oscillating electrons are nothing but an oscillating current, and thus if a large plasmon response is established, resistive heating is a major concern even for small the small resistivities typical of conductors.

This avenue of investigation was motivated by Zawadzka et al \cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors mount a thin gold foil on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the ``grating coupling'' geometry \cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

Every conducting material has an intrinsic plasma frequency given by 
\begin{equation}
  \omega_{P} = \sqrt{\frac{\rho_{e} e^2}{\varepsilon_{0} m^*}} \,\text{,}
\end{equation}
where $\rho_{e}$ is the free-electron density and $m^*$ is the electron effective mass in the material.
And in the most simple (undamped) approximations, the dielectric function of the material may be written in terms of this constant
\begin{equation} \label{eq:dielectric_as_omega}
  \varepsilon(\omega) = 1 - \frac{ \omega_{P}^2 }{ \omega^2 } \,\text{.}
\end{equation}

At the interface of a vacuum and such a simple material, contstraints on the material parameters which may allow surface plasmons, result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $ \cite{cottam_introduction_2004}.
Substituting this result into \ref{eq:dielectric_as_omega} shows that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons propagating along a vacuum interface carry momentum whose magnitude is given by 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

In the grating coupling scheme (shown in \ref{fig:plasmon_schematic}) a laser of wavelength $\lambda$ is incident on a material with a grating of period $d$, making an angle $\theta$ from the surface normal.
A surface plasmon is driven when the sum of the component of the laser wave-vector parallel to the material surface ($k_{Lx}$) and a fourier component ($k_{G} = 2 \pi / d$) of the grating matches the wave-vector of the surface plasmon
\begin{equation}
  k_{SP} = k_{Lx} + n k_{G} \,\text{,}
\end{equation}
where $n$ here is an integer.
This expression reduces to 
\begin{equation} \label{eq:plasmon_condition}
  \sqrt{ \frac{ \varepsilon }{ 1 + \varepsilon } } = \sin \theta + \frac{ n \lambda }{ d } \,\text{,}
\end{equation}
where $\lambda$ is the incident laser wavelength.
For a given grating period, \ref{eq:plasmon_condition} determines the angle of incidence that the laser must make with the surface in order to drive the plasmon.
Alternatively, it may be used to design the periodic surface structure, for example allowing for angles that the microscope column will accept, different driving orders ($n$) and even other Fourier components of the periodic surface structure.

@@

For proof-of-concept purposes the first experiment was run on a commercially available gold foil on glass holographic grating (750 lines/mm).
Like the Kretschmann geometry, a foil on glass grating is vulnerable to heating problems; a plasmon creates a large amount of heat but the glass is a poor heat conductor.
The aforementioned laser (emitting $\sim$200fs pulses of green light at a peak pulse intensity of 30MW/cm$^2$) was used to drive the plasmon.

@@

The input laser angle was slowly changed until the plasmon oscillation is visible.
At the instant that the plasmon started, a bright flash was visible on the scintillator screen and then disappeared.
After this, the grating showed visual signs of damage, both on the surface and in the reflected light pattern.
\ref{fig:grating-damage} shows the results of Scanning Electron Microscopy (SEM) on the grating.
The damage pattern is round and of the size of the laser spot used to drive the plasmon.
The extent of the damage indicates a large temperature increase, which given the other evidence of the test, can be assumed to be from the sudden appearance of a surface plasmon.

There are several conclusions that can be drawn from this experiment.
Most importantly, it shows that a plasmon was driven successfully given the angular dependence and large response, few other processes could have occurred.
The heat generated by the plasmon oscillation destroyed the grating, therefore it validated the decision not to pursue a Kretschmann geometry over fears of heat damage at the high heating rates of the laser-plasmon interaction.
Finally, it suggests that if using a thermally conductive and structurally sound photoemission material and mount for the plasmonic photocathode, PAPE might generate a large number of electrons (as evident by the bright flash).

@@

After the destruction of the gold foil on glass grating, new gratings were designed to have a higher subtrate thermal conductivity.
The second generation grating was constructed from a silicon substract then coated with a gold film.
The thermal conductivity of silicon is approximately 100 times that of glass. %TODO reference needed?

@@

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to iteratively mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$ \cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.
A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is several times thicker than the penetration depth of green light on gold, the laser radiation cannot interact with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?
